---
title: "RV rarefaction curve"
author: "Stephanie Goya"
date: "2023-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Rarefaction curves.
In 2021 there were 69 unique genotypes detected within a total of 883 observations, in 2022 there were 48 unique genotypes with 120 observations.
RVA: 550 observations with 36 unique genotypes in 2021, and 38 observations with 22 unique genotypes in 2022.
RVB: 86 observations with 9 unique genotypes in 2021, and 33 observations with 8 unique genotypes in 2022
RVC: 247 observations with 24 unique genotypes in 2021, and 49 observations with 18 unique genotypes in 2022.

#Rarefaction curves per year of sample collection (Period 1 = 2021, Period 2= 2022):
```{r, rarefaction curve by year of sample collection, echo=TRUE}
library(vegan)
library(readxl)

#Loading dataframe:
RV_dataframe <- read_excel("RV-rarefaction-dataframe.xlsx")
RV_dataframe$MONTH <- substr(RV_dataframe$CollectionDate, 1, 7)
RV.MONTH <- table(RV_dataframe$MONTH, RV_dataframe$RVGenotype)
RVA.df <- subset(RV_dataframe, RV_dataframe$RVSpecie=="A")
RVB.df <- subset(RV_dataframe, RV_dataframe$RVSpecie=="B")
RVC.df <- subset(RV_dataframe, RV_dataframe$RVSpecie=="C")

#Rarefaction curve of all RV genotypes by study period:
RV.Period <- table(RV_dataframe$Period, RV_dataframe$RVGenotype)
rarecurve(as.matrix(RV.Period), step = 1, xlab = "Sample Size", ylab = "All RV genotypes", label = TRUE)

#Analysis by study period and species:
RVA.Period <- table(RVA.df$Period, RVA.df$RVGenotype)
rarecurve(as.matrix(RVA.Period), step = 1, xlab = "Sample Size", ylab = "RVA genotypes", label = TRUE)

#Rarefaction curve of RV-B genotypes by study period 
RVB.Period <- table(RVB.df$Period, RVB.df$RVGenotype)
rarecurve(as.matrix(RVB.Period), step = 1, xlab = "Sample Size", ylab = "RVB genotypes", label = TRUE)

#Rarefaction curve of RV-C genotypes by study period 
RVC.Period <- table(RVC.df$Period, RVC.df$RVGenotype)
rarecurve(as.matrix(RVC.Period), step = 1, xlab = "Sample Size", ylab = "RVC genotypes", label = TRUE)
```

#Rarefaction curves per month of sample collection:
```{r rarefaction curve by month of sample collection, echo=TRUE}
#Analysis per month and species:

RVA.MONTH <- table(RVA.df$MONTH, RVA.df$RVGenotype)
rarecurve(as.matrix(RVA.MONTH), step = 1, xlab = "Sample Size", ylab = "RV-A genotypes", label = TRUE)

RVB.MONTH <- table(RVB.df$MONTH, RVB.df$RVGenotype)
rarecurve(as.matrix(RVB.MONTH), step = 1, xlab = "Sample Size", ylab = "RV-B genotypes", label = TRUE)

RVC.MONTH <- table(RVC.df$MONTH, RVC.df$RVGenotype)
rarecurve(as.matrix(RVC.MONTH), step = 1, xlab = "Sample Size", ylab = "RV-C genotypes", label = TRUE)

```

#Estimation of RV genotype richness. 
Coverage-based rarefaction/extrapolation curve is a measure of the estimated coverage of the sample
```{r extrapolated richness, echo=TRUE}
library("iNEXT")

tryRV <- read.csv("~/Documents/HRV/Community/paper/rarefy/tryRV.csv", row.names=1)
out <- iNEXT(tryRV, q=0, datatype="abundance")
ggiNEXT(out, type=1)

tryRVA <- read.csv("~/Documents/HRV/Community/paper/rarefy/tryRVA.csv", row.names=1)
outA <- iNEXT(tryRVA, q=0, datatype="abundance")
ggiNEXT(outA, type=3)

tryRVB <- read.csv("~/Documents/HRV/Community/paper/rarefy/tryRVB.csv", row.names=1)
outB <- iNEXT(tryRVB, q=0, datatype="abundance")
ggiNEXT(outB, type=3)

tryRVC <- read.csv("~/Documents/HRV/Community/paper/rarefy/tryRVC.csv", row.names=1)
outC <- iNEXT(tryRVC, q=0, datatype="abundance")
ggiNEXT(outC, type=3)

```

```{r richness estimation using Chao1 index, echo=TRUE}
#per month for RV-A
ChaoRichness(tryRVA[,c("X2021.02","X2021.04", "X2021.05", "X2021.06", "X2021.07", "X2022.11", "X2022.12")], conf=0.99)

#per month for RV-B
ChaoRichness(tryRVB[,c("X2021.02","X2021.04", "X2021.05", "X2021.06", "X2021.07", "X2022.11", "X2022.12")], conf=0.99)

#per month for RV-C
ChaoRichness(tryRVC[,c("X2021.02","X2021.04", "X2021.05", "X2021.06", "X2021.07", "X2022.11", "X2022.12")], conf=0.99)

```


