---
title: "RV rarefaction curve"
author: "Stephanie Goya"
date: "2023-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Rarefaction curves.
In 2021 there were 69 unique genotypes detected within a total of 883 observations, in 2022 there were 48 unique genotypes with 120 observations.
RVA: 550 observations with 36 unique genotypes in 2021, and 38 observations with 22 unique genotypes in 2022.
RVB: 86 observations with 9 unique genotypes in 2021, and 33 observations with 8 unique genotypes in 2022
RVC: 247 observations with 24 unique genotypes in 2021, and 49 observations with 18 unique genotypes in 2022.

#Rarefaction curves per year of sample collection (StudyPeriod 1 = 2021, StudyPeriod 2= 2022):
```{r, rarefaction curve by year of sample collection, echo=TRUE}
library(vegan)
library(readr)
library(cowplot)

#Loading dataframe:
RV_dataframe <- read_csv("RV-dataframe_20240130.csv")
RV_dataframe$MONTH <- substr(RV_dataframe$CollectionDate, 1, 7)
RV.MONTH <- table(RV_dataframe$MONTH, RV_dataframe$Genotype)
RVA.df <- subset(RV_dataframe, RV_dataframe$RVSpecie=="A")
RVB.df <- subset(RV_dataframe, RV_dataframe$RVSpecie=="B")
RVC.df <- subset(RV_dataframe, RV_dataframe$RVSpecie=="C")

#Rarefaction curve of all RV genotypes by study StudyPeriod:
RV.StudyPeriod <- table(RV_dataframe$StudyPeriod, RV_dataframe$Genotype)
rarecurve(as.matrix(RV.StudyPeriod), step = 1, xlab = "Sample Size", ylab = "All RV genotypes", label = TRUE)

#Analysis by study StudyPeriod and species:
RVA.StudyPeriod <- table(RVA.df$StudyPeriod, RVA.df$Genotype)
rarecurve(as.matrix(RVA.StudyPeriod), step = 1, xlab = "Sample Size", ylab = "RVA genotypes", label = TRUE)

#Rarefaction curve of RV-B genotypes by study StudyPeriod 
RVB.StudyPeriod <- table(RVB.df$StudyPeriod, RVB.df$Genotype)
rarecurve(as.matrix(RVB.StudyPeriod), step = 1, xlab = "Sample Size", ylab = "RVB genotypes", label = TRUE)

#Rarefaction curve of RV-C genotypes by study StudyPeriod 
RVC.StudyPeriod <- table(RVC.df$StudyPeriod, RVC.df$Genotype)
rarecurve(as.matrix(RVC.StudyPeriod), step = 1, xlab = "Sample Size", ylab = "RVC genotypes", label = TRUE)
```

#Rarefaction curves per month of sample collection:
```{r rarefaction curve by month of sample collection, echo=TRUE}
#Analysis per month and species:

RVA.MONTH <- table(RVA.df$MONTH, RVA.df$Genotype)
rarecurve(as.matrix(RVA.MONTH), step = 1, xlab = "Sample Size", ylab = "RV-A genotypes", label = TRUE)

RVB.MONTH <- table(RVB.df$MONTH, RVB.df$Genotype)
rarecurve(as.matrix(RVB.MONTH), step = 1, xlab = "Sample Size", ylab = "RV-B genotypes", label = TRUE)

RVC.MONTH <- table(RVC.df$MONTH, RVC.df$Genotype)
rarecurve(as.matrix(RVC.MONTH), step = 1, xlab = "Sample Size", ylab = "RV-C genotypes", label = TRUE)

```

#Estimation of RV genotype richness. 
Coverage-based rarefaction/extrapolation curve is a measure of the estimated coverage of the sample
```{r extrapolated richness, echo=TRUE}
library("iNEXT")

numberRV <- read.csv("numberRV.csv", row.names=1)
out <- iNEXT(numberRV, q=0, datatype="abundance")
ggiNEXT(out, type=3)

numberRVA <- read.csv("numberRVA.csv", row.names=1)
outA <- iNEXT(numberRVA, q=0, datatype="abundance")
ggiNEXT(outA, type=3)

numberRVB <- read.csv("numberRVB.csv", row.names=1)
outB <- iNEXT(numberRVB, q=0, datatype="abundance")
ggiNEXT(outB, type=3)

numberRVC <- read.csv("numberRVC.csv", row.names=1)
outC <- iNEXT(numberRVC, q=0, datatype="abundance")
ggiNEXT(outC, type=3)

```

```{r richness estimation using Chao1 index, echo=TRUE}
#per month for RV-A
ChaoRichness(numberRVA[,c("X2021.02","X2021.04", "X2021.05", "X2021.06", "X2021.07", "X2022.11", "X2022.12")], conf=0.99)

#per month for RV-B
ChaoRichness(numberRVB[,c("X2021.02","X2021.04", "X2021.05", "X2021.06", "X2021.07", "X2022.11", "X2022.12")], conf=0.99)

#per month for RV-C
ChaoRichness(numberRVC[,c("X2021.02","X2021.04", "X2021.05", "X2021.06", "X2021.07", "X2022.11", "X2022.12")], conf=0.99)

```


